---
title: "README"
author: "Ed Rogers"
date: "05/23/2015"
output: pdf_document
---

# EPADataAnalysis
Course Project 2 for Exploratory Data Analysis

The 6 included scripts (plot1.R, plot2.R, etc.) generate plots to answer the six questions associated with Course Project 2 for the Exploratory Data Analysis course. Each script should be run in the directory in which the EPA National Emissions Inventory has been unzipped.

##Question 1: Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.
```{r,echo=FALSE}
suppressPackageStartupMessages(library(dplyr,quietly = TRUE))

# This first line will likely take a few seconds. Be patient!
NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")

yearlyTotalEmissions <- NEI %>% group_by(year) %>% summarize("Yearly Total Emissions" = sum(Emissions))

par(mar = c(3,6,4,2)+0.1)
par(mgp = c(4,1,0))
plot(yearlyTotalEmissions$year,yearlyTotalEmissions$"Yearly Total Emissions",
     xlab = "",
     ylab = expression('Tons of PM'[2.5]),
     las=1,
     main=expression("United States Total PM"[2.5]*" Emissions by Year"))
abline(lm(yearlyTotalEmissions$"Yearly Total Emissions" ~ yearlyTotalEmissions$year))
```

------

##Question 2: Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008?

```{r,echo=FALSE}
baltimoreCityEmissions <- NEI[NEI$fips =="24510",c("year","Emissions")]
yearlyBCEmissions <- baltimoreCityEmissions %>% group_by(year) %>% summarize("City Total Emissions" = sum(Emissions))

plot(yearlyBCEmissions$year,yearlyBCEmissions$"City Total Emissions",
     xlab = "",
     ylab = expression('Tons of PM'[2.5]),
     las=1,
     main=expression("Baltimore City Total PM"[2.5]*" Emissions by Year"))
abline(lm(yearlyBCEmissions$"City Total Emissions" ~ yearlyBCEmissions$year))
```

------

##Question 3: Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008?

```{r,echo=FALSE}
library(ggplot2)
baltimoreCityEmissions <- NEI[NEI$fips =="24510",c("year","type","Emissions")]
yearlyBCEmissions <- baltimoreCityEmissions %>% group_by(year,type) %>% summarize("Emissions" = sum(Emissions))

p = ggplot(data = yearlyBCEmissions,aes(year,Emissions,colour=yearlyBCEmissions$type))+
  geom_point(size=4)+
  geom_line()+
  xlab("")+
  ylab(expression("Tons of PM"[2.5]))+
  ggtitle(expression("Baltimore City PM"[2.5]*" Emissions by Year"))+
  theme(legend.title=element_blank())
print(p)
```

------

##Question 4: Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?

```{r,echo=FALSE}
# What SCC codes are associated with coal combustion?
coalSectors <- grep("coal",levels(SCC$EI.Sector),ignore.case = TRUE,value=TRUE)
coalCodes <- SCC[ SCC$EI.Sector %in% coalSectors, "SCC" ]
#What are the associated emissions?
coalEmissions <- NEI[ NEI$SCC %in% coalCodes,c("year","SCC","Emissions")]
yearlyCoalEmissions <- coalEmissions %>% group_by(year) %>% summarize("Emissions" = sum(Emissions))

p = ggplot(data = yearlyCoalEmissions,aes(year,Emissions))+
  geom_point(size=4)+
  xlab("")+
  ylab(expression("Tons of PM"[2.5]))+
  ggtitle(expression("United States PM"[2.5]*" Emissions from Coal Combustion by Year"))+
  theme(legend.title=element_blank())+
  stat_smooth(method="lm", se=FALSE)
print(p)
```

------

## Question 5: How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?

```{r,echo=FALSE}
# What SCC codes are associated with motor vehicles?
allVehicles   <- grep("vehicle",levels(SCC$Short.Name),ignore.case = TRUE,value=TRUE)
# Note: Vehicle Cooking is probably not "motor vehicle source. 
# Exclude "vehicle cooking as a source of codes
motorVehicles <- grep("vehicle cooking",allVehicles,ignore.case = TRUE,invert=TRUE,value = TRUE)
# Use the codes to select rows and relevant columns from NEI
motorVehicleCodes <- SCC[ SCC$Short.Name %in% motorVehicles, "SCC" ]
mVehicleEmissions <- NEI[ NEI$SCC %in% motorVehicleCodes,
                              c("year","SCC","fips","Emissions")]
# Subset to Baltimore City
bCityMotorVehicleEmissions <- mVehicleEmissions[ mVehicleEmissions$fips==24510,]
# Sum the emissions by year
yearlyEmissions <- bCityMotorVehicleEmissions %>% group_by(year) %>% summarize("Emissions" = sum(Emissions))

p = ggplot(data = yearlyEmissions,aes(year,Emissions))+
  geom_point(size=4)+
  xlab("")+
  ylab(expression("Tons of PM"[2.5]))+
  ggtitle(expression("Baltimore City PM"[2.5]*" Emissions from Motor Vehicles by Year"))+
  theme(legend.title=element_blank())+
  stat_smooth(method="lm", se=FALSE)
print(p)
```

------

##Question 6: Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?

```{r,echo=FALSE,fig.width=7, fig.height=6,out.height="10in",out.width="7in"}
suppressPackageStartupMessages(library(gridExtra,quietly = TRUE))
library(scales)

# What SCC codes are associated with motor vehicles?
allVehicles   <- grep("vehicle",levels(SCC$Short.Name),ignore.case = TRUE,value=TRUE)
# Note: Vehicle Cooking is probably not "motor vehicle source. 
# Exclude "vehicle cooking as a source of codes
motorVehicles <- grep("vehicle cooking",allVehicles,ignore.case = TRUE,invert=TRUE,value = TRUE)
# Use the codes to select rows and relevant columns from NEI
motorVehicleCodes <- SCC[ SCC$Short.Name %in% motorVehicles, "SCC" ]
mVehicleEmissions <- NEI[ NEI$SCC %in% motorVehicleCodes,
                          c("year","SCC","fips","Emissions")]
# Subset to Baltimore City & Los Angeles County
cityMotorVehicleEmissions <- mVehicleEmissions[ mVehicleEmissions$fips %in% c("24510","06037"),]
cityMotorVehicleEmissions$fips <- factor(cityMotorVehicleEmissions$fips,labels=c("Los Angeles County","Baltimore City"))
# Sum the emissions by year
yearlyEmissions <- cityMotorVehicleEmissions %>% group_by(year,fips) %>% summarize("Emissions" = sum(Emissions))
# Now scale them to the 1999 values, to compare relative change
yearlyEmissionsPercent <- ungroup(yearlyEmissions) %>% group_by(fips) %>% mutate(Emissions = Emissions/first(Emissions))

absolutePlot = ggplot(data = yearlyEmissions,aes(year,Emissions,group=yearlyEmissions$fips,color=yearlyEmissions$fips))+
  geom_point(size=4)+
  geom_line()+
  xlab("")+
  ylab(expression("Tons of PM"[2.5]))+
  ggtitle(expression(atop("PM"[2.5]*" Emissions from Motor Vehicles by Year", atop(italic("Total Yearly Emissions Between Cities"), ""))))+
  theme(legend.title=element_blank())
percentPlot = ggplot(data = yearlyEmissionsPercent,aes(year,Emissions,group=yearlyEmissionsPercent$fips,color=yearlyEmissionsPercent$fips,fill=yearlyEmissionsPercent$fips))+
  geom_point(size=4)+
  geom_line()+
  xlab("")+
  ylab(expression("Percentage PM"[2.5]))+
  ggtitle(expression(atop("PM"[2.5]*" Emissions from Motor Vehicles by Year", atop(italic("Percentage of Emissions Relative to 1999 Measurements"), ""))))+
  theme(legend.title=element_blank())+
  scale_y_continuous(labels=percent_format(),limits=c(0,1))
grid.arrange(absolutePlot,percentPlot,ncol=1)
```